---
title: "Data Cleaning and analysis"
author: "Chris Bocz"
date: "2025-04-18"
output: html_document
---

## Experimental Goals

This study examines how cover crop planting and termination timing affect soil moisture and temperature in cotton fields in Georgia. Results may guide growers on whether to adopt cover crops, and particularly whether to adopt the "Planting Green" practice of planting cotton into living cereal rye terminated just after planting, rather than 14 days before. Planting Green may enhance soil water retention, improve planting conditions, and suppress weeds, but its effects in Georgia remain untested.

## Experimental Design

The experiment leveraged four existing complete randomized block trials: two in Midville, Ga (sandy soil) and two in Watkinsville, Ga (clay soil); each location had one trial planted to cereal rye late in October 2023, and one planted to cereal rye in late November 2023. Each trial included three treatments with four replications:

1.   No cover (nc) – no rye planted.

2.   Plant brown (pb) – rye terminated 14 days before cotton planting.

3.   Plant green (pg) – rye terminated 1–3 days after cotton planting.

Due to funding limits, sensors were installed in one replication per trial. In each treatment, two sensors (5 cm and 20 cm depths) measured six variables every minute: volumetric water content (vwc), electrical conductivity (ec), temperature (t), signal attenuation (p), propagation time (pa), and volumetric reflectivity (vr). The latter three variables were used to calculate vwc and ec. Daily and hourly min, max, and average values were recorded.

This analysis focuses on vwc and soil temperature, examining patterns across cover crop treatments, planting/termination timings, soil types, and depths. It is important to note that that there is not enough replication in these data to find group means significant, therefore this analysis will only investigate potential patterns for further study.

\

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rlang)  # for .data pronoun
```

## Load Data

```{r list, echo=TRUE, warning=FALSE, message=FALSE}
# List Raw Data files in the working directory (raw files from data-loggers are in .dat format).  Create a list of all files, then 2 seperate lists for files from Watkinsville vs Midville experiments.

file_paths <- list.files(
  path = "Data/",
  pattern = "\\.dat$",
  full.names = TRUE
)
mdv_files <- file_paths[grepl("Midville", file_paths, ignore.case = TRUE)]
wat_files <- file_paths[grepl("Watkinsville", file_paths, ignore.case = TRUE)]

#Create data-frames, 1 for Midville, 1 for Watkinsville.  Add filename as a column, for extracting metadata later.  Skip first row of observations; standard for data files from our Campbell Scientific loggers 

mdv_df <- map_dfr(mdv_files, function(file) {
  df <- read_csv(file, skip = 1)  # Read file, skipping the first row
  df %>% mutate(source_file = basename(file))  # Add filename as a column
})

wat_df <- map_dfr(wat_files, function(file) {
  df <- read_csv(file, skip = 1)  # Read file, skipping the first row
  df %>% mutate(source_file = basename(file))  # Add filename as a column
})

```

## Data Cleaning

```{r Midville, echo=TRUE, warning=FALSE}
#Cleaning Data from the Data-Loggers from Midville experiments
# Extract the units row
mdv_units_row <- mdv_df[1, ]

# Remove the units row and next row from the data
mdv_cleaned_data1 <- mdv_df[-c(1,2),]

# Combine column names and units
new_colnames <- paste0(names(mdv_cleaned_data1), " (", unlist(mdv_units_row), ")")
names(mdv_cleaned_data1) <- new_colnames
mdv_cleaned_data1 <- mdv_cleaned_data1 %>%
  rename(`source file` = `source_file (Cotton Commission Midville Early_Daily.dat)`)

# Remove un-needed columns
mdv_cleaned_data2 <- mdv_cleaned_data1 %>%
   select(-`RECORD (RN)`, -`BattV_Min (Volts)`, -`PTemp_C_Avg (Deg C)`, -`BattV_Avg (NA)`)

# Remove duplicate/missing data, convert column names to lower-case, replace spaces with underscores.  This is general practice to standardize column names and make them universally readable to Tinyverse
mdv_cleaned_data3 <- mdv_cleaned_data2 %>%
  na.omit() %>%  # Remove rows with missing values
  distinct() %>%  # Remove duplicate rows
  rename_all(tolower) %>%  # Convert column names to lowercase
  rename_all(~gsub(" ", "_", .))  # Replace spaces with underscores

# Convert the original timestamp column to date-time format and rename it, remove timestamp_(ts) column
mdv_cleaned_data4 <- mdv_cleaned_data3 %>%
  mutate(timestamp = ymd_hms(`timestamp_(ts)`)) %>%
  select(-`timestamp_(ts)`, everything())

#Remove NA's introduced in the last step:
mdv_cleaned_data5 <- na.omit(mdv_cleaned_data4)

## Filter the dataset to remove observations before sensors were installed and after sensors were pulled (October 11, 2024 in Midville, November 11, 2024 in Watkinsville)  
mdv_cleaned_data6 <- mdv_cleaned_data5 %>%
  filter(timestamp <= as.POSIXct("2024-10-11 23:59:59"))

# Pivot to long data, make sensor depth, cover-crop treatment and "stat" into variables. Stat: loggers took min, max, and avg data for each observation, the "stat" column tells us which of those is being reported.  Also pull metadata from filenames (sourcefile column), create cc_plant_time (early vs late planted cover crop), and dara_freq (whether observation is hourly or daily)

mdv_cleaned_data7 <- mdv_cleaned_data6 %>%
  pivot_longer(
    cols = -c(timestamp, source_file),
    names_to = c("measure", "treatment", "depth", "stat", "units"),
    names_sep = "_"
  ) %>%
  mutate(
    # Extract metadata from source_file
    cc_plant_time = str_extract(source_file, "Early|Late") %>% tolower(),
    data_freq = str_extract(source_file, "Daily|Hourly") %>% tolower(),
  
    # Handle measure units
    measure_units = case_when(
     measure == "t" & units == "(deg" ~ "t_(deg_c)",
      TRUE ~ paste0(measure, "_", units)
    ),
    value = as.numeric(value)
  )%>%
  select(-measure, -units) %>%
  filter(!is.na(measure_units)) #%>%
  # Rename and format columns
  #rename_with(~ tolower(gsub(" ", "_", .x)))  # Applies to all columns

  #select(-measure, -units) %>%
  #filter(!is.na(measure_units))  # Remove rows where measure_units is NA

#Remove source_file column
mdv_cleaned_data8 <- mdv_cleaned_data7 %>% 
  select(-source_file)

#Pivot the data so that measure_units becomes columns, means a seperate column for each collected-data variable
mdv_cleaned_data9 <- mdv_cleaned_data8 %>%
  pivot_wider(names_from = measure_units, values_from = value) %>%
    mutate(
      treatment = as.factor(treatment),
      depth = as.factor(depth),
      stat = as.factor(stat),
      cc_plant_time = as.factor(cc_plant_time),
      data_freq = as.factor(data_freq)
      ) %>% 
  select(-timestamp_NA)#remove uneeded column created by pivot_wider

# Remove observations where `vwc_(m^3/m^3)` <= 0.0000 or `ec_(ds/m)` < 0.  This should remove observations where loggers were on but not installed.  Remove NAs.
mdv_cleaned_data10 <- mdv_cleaned_data9 %>%
  filter(!(`vwc_(m^3/m^3)` <= 0 | `ec_(ds/m)` < 0)) %>% 
  na.omit()

# Create separate tables for all 'stat' * 'data_freq' combinations

mdv_daily_avg <- mdv_cleaned_data10 %>%
  filter(stat == "avg", data_freq == "daily")

mdv_daily_min <- mdv_cleaned_data10 %>%
  filter(stat == "min", data_freq == "daily")

mdv_daily_max <- mdv_cleaned_data10 %>%
  filter(stat == "max", data_freq == "daily")

mdv_hourly_avg <- mdv_cleaned_data10 %>%
  filter(stat == "avg", data_freq == "hourly")

mdv_hourly_min <- mdv_cleaned_data10 %>%
  filter(stat == "min", data_freq == "hourly")

mdv_hourly_max <- mdv_cleaned_data10 %>%
  filter(stat == "max", data_freq == "hourly")

```

```{r Watkinsville, echo=TRUE, warning=FALSE}
#Cleaning Data from the Data-Loggers from Watkinsville experiments
# Extract the units row
wat_units_row <- wat_df[1, ]

# Remove the units row and next row from the data
wat_cleaned_data1 <- wat_df[-c(1,2),]

# Combine column names and units
new_colnames <- paste0(names(wat_cleaned_data1), " (", unlist(wat_units_row), ")")
names(wat_cleaned_data1) <- new_colnames
wat_cleaned_data1 <- wat_cleaned_data1 %>%
  rename(`source file` = `source_file (Cotton Commission Watkinsville Early_Daily.dat)`)

# Remove un-needed columns
wat_cleaned_data2 <- wat_cleaned_data1 %>%
   select(-`RECORD (RN)`, -`BattV_Min (Volts)`, -`PTemp_C_Avg (Deg C)`, -`BattV_Avg (NA)`)

# Remove duplicate/missing data, convert column names to lower-case, replace spaces with underscores.  This is good-practice to standardize column names and make them universally readable to Tinyverse
wat_cleaned_data3 <- wat_cleaned_data2 %>%
  na.omit() %>%  # Remove rows with missing values
  distinct() %>%  # Remove duplicate rows
  rename_all(tolower) %>%  # Convert column names to lowercase
  rename_all(~gsub(" ", "_", .))  # Replace spaces with underscores

# Convert the original timestamp column to date-time format and rename it, remove timestamp_(ts) column
wat_cleaned_data4 <- wat_cleaned_data3 %>%
  mutate(timestamp = ymd_hms(`timestamp_(ts)`)) %>%
  select(-`timestamp_(ts)`, everything())

#Remove NA's introduced in the last step:
wat_cleaned_data5 <- na.omit(wat_cleaned_data4)

## Filter the dataset to remove observations before sensors were installed and after sensors were pulled (October 11, 2024 in Midville, November 11, 2024 in Watkinsville)  
wat_cleaned_data6 <- wat_cleaned_data5 %>%
  filter(timestamp <= as.POSIXct("2024-11-11 23:59:59"))

# Pivot to long data, make sensor depth, cover-crop treatment and "stat" into variables. Stat: loggers took min, max, and avg data for each observation, the "stat" column tells us which of those is being reported.  Also pull metadata from filenames (sourcefile column), create cc_plant_time (early vs late planted cover crop), and dara_freq (whether observation is hourly or daily)

wat_cleaned_data7 <- wat_cleaned_data6 %>%
  pivot_longer(
    cols = -c(timestamp, source_file),
    names_to = c("measure", "treatment", "depth", "stat", "units"),
    names_sep = "_"
  ) %>%
  mutate(
    # Extract metadata from source_file
    cc_plant_time = str_extract(source_file, "Early|Late") %>% tolower(),
    data_freq = str_extract(source_file, "Daily|Hourly") %>% tolower(),
  
    # Handle measure units
    measure_units = case_when(
     measure == "t" & units == "(deg" ~ "t_(deg_c)",
      TRUE ~ paste0(measure, "_", units)
    ),
    value = as.numeric(value)
  )%>%
  select(-measure, -units) %>%
  filter(!is.na(measure_units))
  # Rename and format columns
  #rename_with(~ tolower(gsub(" ", "_", .x)))  # Applies to all columns

  #select(-measure, -units) %>%
  #filter(!is.na(measure_units))  # Remove rows where measure_units is NA

#Change treatment, depth, stat, cc_plant_time, and data_freq columns to factors, remove source_file column
wat_cleaned_data8 <- wat_cleaned_data7 %>%
  mutate(
      treatment = as.factor(treatment),
      depth = as.factor(depth),
      stat = as.factor(stat),
      cc_plant_time = as.factor(cc_plant_time),
      data_freq = as.factor(data_freq)
      ) %>% 
  select(-source_file)

#Pivot the data so that measure_units becomes columns, means a seperate column for each collected-data variable
wat_cleaned_data9 <- wat_cleaned_data8 %>%
  pivot_wider(names_from = measure_units, values_from = value) %>% 
  select(-timestamp_NA)#remove uneeded column created by pivot_wider

# Remove observations where `vwc_(m^3/m^3)` <= 0.0000 or `ec_(ds/m)` < 0.  This should remove observations where loggers were on but not installed.  Remove NAs.
wat_cleaned_data10 <- wat_cleaned_data9 %>%
  filter(!(`vwc_(m^3/m^3)` <= 0 | `ec_(ds/m)` < 0)) %>% 
  na.omit()

# Create separate tables for all 'stat' * 'data_freq' combinations

wat_daily_avg <- wat_cleaned_data10 %>%
  filter(stat == "avg", data_freq == "daily")

wat_daily_min <- wat_cleaned_data10 %>%
  filter(stat == "min", data_freq == "daily")

wat_daily_max <- wat_cleaned_data10 %>%
  filter(stat == "max", data_freq == "daily")

wat_hourly_avg <- wat_cleaned_data10 %>%
  filter(stat == "avg", data_freq == "hourly")

wat_hourly_min <- wat_cleaned_data10 %>%
  filter(stat == "min", data_freq == "hourly")

wat_hourly_max <- wat_cleaned_data10 %>%
  filter(stat == "max", data_freq == "hourly")


```

## Exploratory Data Analysis

This will be the only data analysis phase, because there are not enough replications to give the degrees of freedom necessary to determine significant trends or differences in group means. This analysis will explore the effects of cover crop planting time, sensor depth, and cover crop treatment on response variables: soil volumetric water content (vwc\_(m^3/m^3)), electrical conductivity (ec\_(ds/m)), and temperature (t\_(deg_c)). However users can pick response variables to investigate by redefining "response_vars" in the "objects" chunk.

```{r objects required for looping functions, echo=TRUE, warning=FALSE}
# List your cleaned data frames and give them friendly names for plotting
data_list <- list(
  wat = wat_cleaned_data10,
  mdv = mdv_cleaned_data10
)

# List combinations of 'stat' * 'data_freq'
combinations <- wat_cleaned_data10 %>%
  distinct(stat, data_freq) %>% 
  filter(data_freq == "hourly")

# Define response variables in a coding-friendly way, this makes it easier to change variables if needed
response_vars <- c("vwc_(m^3/m^3)", "ec_(ds/m)", "t_(deg_c)")

```

```{r histogram, eval=FALSE, include=FALSE}
##Using `vwc_(m^3/m^3)' as response variable, create histograms for all permutations of 'stat' * 'data_freq' at both locations

# Create plots, looping through all possible combinations of 'stat'*'data_freq'

for (loc in names(data_list)) {
  data <- data_list[[loc]]
  
  for (i in 1:nrow(combinations)) {
    stat_val <- combinations$stat[i]
    freq_val <- combinations$data_freq[i]
    
    subset_data <- data %>%
      filter(stat == stat_val, data_freq == freq_val)
    
    for (resp in response_vars) {
      p <- ggplot(subset_data, aes(x = .data[[resp]], fill = treatment)) +
        geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
        labs(
          title = paste(loc, "|", resp, "|", stat_val, "|", freq_val),
          x = resp,
          y = "Count"
        ) +
        theme_minimal()
      
      print(p)
    }
  }
}


```

```{r boxplots, eval=FALSE, include=FALSE}

for (loc in names(data_list)) {
  data <- data_list[[loc]]
  
  # Ensure depth is a factor with the desired order
  data <- data %>%
    mutate(depth = factor(depth, levels = c("5cm", "20cm")))
  
  for (i in 1:nrow(combinations)) {
    stat_val <- combinations$stat[i]
    freq_val <- combinations$data_freq[i]
    
    subset_data <- data %>%
      filter(stat == stat_val, data_freq == freq_val)
    
    for (resp in response_vars) {
      p <- ggplot(subset_data, aes(x = treatment, y = .data[[resp]], fill = depth)) +
        geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2) +
        labs(
          title = paste(loc, "|", resp, "|", stat_val, "|", freq_val),
          x = "Treatment",
          y = resp
        ) +
        theme_minimal()
      
      print(p)
    }
  }
}

for (loc in names(data_list)) {
  data <- data_list[[loc]]
  
  # Ensure depth is a factor with the desired order
  
  for (i in 1:nrow(combinations)) {
    stat_val <- combinations$stat[i]
    freq_val <- combinations$data_freq[i]
    
    subset_data <- data %>%
      filter(stat == stat_val, data_freq == freq_val)
    
    for (resp in response_vars) {
      p <- ggplot(subset_data, aes(x = treatment, y = .data[[resp]], fill = cc_plant_time)) +
        geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2) +
        labs(
          title = paste(loc, "|", resp, "|", stat_val, "|", freq_val),
          x = "Treatment",
          y = resp
        ) +
        theme_minimal()
      
      print(p)
    }
  }
}



```

```{r relationships over time, eval=FALSE, include=FALSE}
#create plots which show response vars overtime for all combinations
#for (loc in names(data_list)) {
  #data <- data_list[[loc]]
  
  #data <- data %>%
   # mutate(depth = factor(depth, levels = c("5cm", "20cm")))# Ensure depth is a factor with the desired order
  
  #for (i in 1:nrow(combinations)) {
    #stat_val <- combinations$stat[i]
    #freq_val <- combinations$data_freq[i]
    
    #subset_data <- data %>%
      #filter(stat == stat_val, data_freq == freq_val)
    
    #for (resp in response_vars) {
      #p <- ggplot(subset_data, aes(x = timestamp, y = .data[[resp]], color = treatment, linetype = depth)) +
        #geom_line() +
        #labs(
          #title = paste(loc, "|", resp, "|", stat_val, "|", freq_val),
          #x = "Date",
          #y = resp
        #) +
        #theme_minimal()
      
      #print(p)
    #}
  #}
#}
#same plots, but with smoothing function to create trendline to reduce noise
for (loc in names(data_list)) {
  data <- data_list[[loc]] %>%
    mutate(depth = factor(depth, levels = c("5cm", "20cm")))
  
  for (i in 1:nrow(combinations)) {
    stat_val <- combinations$stat[i]
    freq_val <- combinations$data_freq[i]
    
    subset_data <- data %>%
      filter(stat == stat_val, data_freq == freq_val)
    
    for (resp in response_vars) {
      p <- ggplot(subset_data, aes(x = timestamp, y = .data[[resp]], color = treatment, linetype = depth)) +
        geom_smooth(span = 0.3, se = FALSE) +  # Smoothing line only
        labs(
          title = paste(loc, "| Smoothed", resp, "|", stat_val, "|", freq_val),
          x = "Date",
          y = resp
        ) +
        theme_minimal()
      
      print(p)
    }
  }
}




```

```{r Plots, echo=TRUE, warning=FALSE}
#boxplots

for (loc in names(data_list)) {
  data <- data_list[[loc]]
  
  data <- data %>%
    mutate(depth = factor(depth, levels = c("5cm", "20cm")))
  
  for (i in 1:nrow(combinations)) {
    stat_val <- combinations$stat[i]
    freq_val <- combinations$data_freq[i]
    
    subset_data <- data %>%
      filter(stat == stat_val, data_freq == freq_val)
    
    for (resp in response_vars) {
      p <- ggplot(subset_data, aes(x = treatment, y = .data[[resp]], fill = cc_plant_time)) +
        geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2) +
        facet_wrap(~depth)+
        labs(
          title = paste(loc, "|", resp, "|", stat_val, "|", freq_val),
          x = "Treatment",
          y = resp
        ) +
        theme_minimal()
      
      print(p)
    }
  }
}


#lineplots
for (loc in names(data_list)) {
  data <- data_list[[loc]] %>%
    mutate(depth = factor(depth, levels = c("5cm", "20cm")))
  
  for (i in 1:nrow(combinations)) {
    stat_val <- combinations$stat[i]
    freq_val <- combinations$data_freq[i]
    
    subset_data <- data %>%
      filter(stat == stat_val, data_freq == freq_val)
    
    for (resp in response_vars) {
      p <- ggplot(subset_data, aes(x = timestamp, y = .data[[resp]], color = treatment, linetype = cc_plant_time)) +
        geom_smooth(span = 0.3, se = FALSE) +  # Smoothing line only
        labs(
          title = paste(loc, "| Smoothed", resp, "|", stat_val, "|", freq_val),
          x = "Date",
          y = resp
        ) +
        facet_wrap(~depth)+
        theme_minimal()
      
      print(p)
    }
  }
}


```

```{}
```

Preliminary Conclusions, and Further Work

This analysis suggests that early planting of the cover crop (cc_plant_time early) increased volumetric water content and electrical conductivity throughout the season. This supports the hypothesis that the increased soil organic matter from an early-planted cover-crop will increase soil water infiltration and retention.

There was some evidence for higher soil temperatures early in the season in treatments with no cover-crop. This temperature difference disappeared near the beginning of August. This pattern is likely due to the lack of cover crop residue shading the soil early in the season, leading to higher soil temperatures which then cooled as the cotton crop formed a canopy, effectively shading the soil as in the other treatments. \
\
Further analysis of these data could further investigate the soil temperature difference specifically before August.\
\
Further experiments in this area should focus on creating enough replication to determine significant differences in group means, and significant trends.
